\section{Software and Verification}

\subsection{Why testbenches matter}

A CPU can appear to run simple code while still having subtle bugs in branch offsets, sign extension, load alignment, forwarding, or trap priority. Testbenches catch these by running known programs and checking architectural results.

The repository includes step-by-step docs and testbenches that validate individual blocks and integrated cores.

\subsection{What to test per block}

\begin{longtable}{|>{\RaggedRight\arraybackslash}p{0.38\textwidth}|>{\RaggedRight\arraybackslash}p{0.52\textwidth}|}
\hline
\textbf{Block} & \textbf{Important tests} \\
\hline
\endfirsthead
\hline
\textbf{Block} & \textbf{Important tests} \\
\hline
\endhead
ALU & Signed/unsigned comparisons, shifts, multiply high variants, division edge cases. \\
\hline
Immediate generator & Negative immediates, B/J offsets, U immediates. \\
\hline
Register file & x0 hardwired zero, simultaneous reads, write enable. \\
\hline
Control/decode & Every opcode family, RV32M funct7 detection, load/store/branch controls. \\
\hline
Branch logic & Signed and unsigned branch conditions, negative branch offsets. \\
\hline
Load/store & Byte/halfword/word, sign/zero extension, unaligned behavior if defined. \\
\hline
CSR/trap & \texttt{ecall}, \texttt{ebreak}, illegal instruction, timer interrupt, \texttt{mret}. \\
\hline
Pipeline & Forwarding, load-use stall, branch flush, write-back hazards. \\
\hline
Branch predictor & Correctness with mispredictions, performance counters, BTB aliasing behavior. \\
\hline
MMU & Valid translations, invalid PTEs, permission faults, superpages, user/M-mode bypass. \\
\hline
Atomics & AMO old/new values, LR/SC success/failure, reservation clearing. \\
\hline
\end{longtable}

\subsection{Software demos}

The \texttt{sw/} directory contains generated hex files and software experiments. A typical flow is:

\begin{lstlisting}[language={}]
compile/link bare-metal C or assembly
convert ELF/program to hex
load hex into instruction memory
run simulation or FPGA system
observe memory/UART/testbench outputs
\end{lstlisting}

This makes the CPU verifiable at the architectural level, not only as isolated HDL blocks.

\bigskip
